\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-22 analysis/paper/paper.Rmd
final version

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function and therefore needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through four different case studies.
  | *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, "\\proglang{R}", data visualization]
  plain:     [survival analysis, absolute risk, R, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

echo_plot_code <- TRUE

paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Using the framework of logistic regression for analyzing time-to-event data opens the door to an extensive array of powerful modeling tools. Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function. Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.

@@

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder $\mathrm{past}$ to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, their sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base (e.g. the sum of all individual follow-up times). More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.

@@

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and solve the following problem:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

@@

$$ CI_j(t) = \int_0^t f_j(u)du.$$
The integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration. This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.

<!--We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}. The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.-->

@@

\subsection{Different functional forms of time}

@@

```{r, results='hide', eval = FALSE, echo = FALSE}
# par(mfrow = c(2, 2))
# lapply(fits, function(i) {
#   casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
#   casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
# })
```

```{r erspc-casebase-exponential, eval = FALSE, echo = FALSE}
# # Equivalence between casebase and AFT
# # exponential -------------------------------------------------------------
# 
# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
#                                                   data = ERSPC, 
#                                                   ratio = 100)
# aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
#                            data = ERSPC, dist = "exp")
# 
# -coef(aft_exponential)
# coef(casebase_exponential)
# summary(casebase_exponential)
# exp(coef(casebase_exponential)[2])
# exp(confint(casebase_exponential)[2, ])
# 
# # weibull -----------------------------------------------------------------
# 
# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
#                                            data = ERSPC, 
#                                            ratio = 100)
# aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
#                        data = ERSPC, dist = "weibull")
# 
# -coef(aft_weibull)/aft_weibull$scale
# coef(casebase_weibull)
```

<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any). -->
<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->

@@

```{r erspc-95ci, echo = FALSE, eval = FALSE}
# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
# CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
# CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
# CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])
# 
# table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
# colnames(table_ci) <- c("HR", "LB", "UB")
# 
# data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>%
#   bind_cols(as.data.frame(table_ci)) %>%
#   mutate(
#     CI = paste0(
#       "(", round(LB, 2), ", ",
#       round(UB, 2), ")"
#     ),
#     Width = UB - LB
#   ) %>%
#   dplyr::select(-LB, -UB) %>%
#   knitr::kable(format = "latex", booktabs = TRUE) %>%
#   kable_styling()
```

\subsection{Time-dependent hazard ratios}

@@

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

@@

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}

# library(casebase)
# library(splines)
# data("ERSPC")
# ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
#                        levels = c(0,1), 
#                        labels = c("Control group", "Screening group"))
# set.seed(12345)
# fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time,2)*ScrArm, data = ERSPC)
# summary(fit_inter)

# library(visreg)
# plot(fit_inter,
#      type = "hazard",
#      hazard.params = list(xvar = "Follow.Up.Time",
#                           by = "ScrArm",
#                           alpha = 0.05,
#                           ylab = "Hazard"))

# reference category
#dev.off()
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit_inter,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = T)
# abline(a=1,b=0, lty = 2)
axis(2, at = seq(0,1.50, by=0.25), las = 2)
axis(1, at = seq(min(new_time),max(new_time), by=1))
for( i in seq(0,1.50, by=0.25)) {
  abline(h = i, lty = 1, col = "lightgrey")
}
for( i in seq(min(new_time),max(new_time), by=1)) {
  abline(v = i, lty = 1, col = "lightgrey")
}
for(i in seq_along(seq(0, 0.75, by = 0.25))) {
  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
}
lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
```

@@

\newpage

@@

In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.

@@

The data contains information on 177 patients who received a stem-cell transplant for acute leukemia. The event of interest is relapse, but other competing causes (e.g. death, progression, graft failure, graft-versus-host disease) were also recorded. Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age. <!--A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}. We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.-->

```{r echo = FALSE, eval = FALSE}
tibble::tribble(
  ~Variable, ~Description, ~`Statistical summary`,
  "\\code{Sex}", "Sex", "M=Male (100)",
  "", "", "F=Female (77)",
  "\\code{D}", "Disease", "ALL (73)",
  "", "", "AML (104)",
  "\\code{Phase}", "Phase", "CR1 (47)",
  "", "", "CR2 (45)",
  "", "", "CR3 (12)",
  "", "", "Relapse (73)",
  "\\code{Source}", "Type of transplant", "BM+PB (21)",
  "", "", "PB (156)",
  "\\code{Age}", "Age of patient (years)", "4--62",
  "", "", "30.47 (13.04)",
  "\\code{ftime}", "Failure time (months)", "0.13--131.77",
  "", "", "20.28 (30.78)",
  "\\code{Status}", "Status indicator", "0=censored (46)",
  "", "", "1=relapse (56)",
  "", "", "2=competing event (75)"
) %>%
  knitr::kable(
    format = "latex", booktabs = TRUE,
    caption = "Baseline characteristics of patients in the stem-cell transplant study.",
    label = "table1bmtcrr",
    escape = FALSE
  ) %>%
  kable_styling()
```

First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B). In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}

# fill_cols <- colorspace::sequential_hcl(n = 3, palette = "Viridis")
# 
# (fill_colors <- c("Case series" = fill_cols[1],
#                  "Competing event" = fill_cols[3],
#                  "Base series" = fill_cols[2]))
# 
# color_cols <- colorspace::darken(col = fill_cols, amount = 0.3)
# 
# (color_colors <- c("Case series" = color_cols[1],
#                    "Competing event" = color_cols[3],
#                    "Base series" = color_cols[2]))
# 
# pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
# 
# pt_object_comp <- bmtcrr %>%
#   mutate(Status = case_when(
#     Status == 1 ~ 2L,
#     Status == 2 ~ 1L,
#     TRUE ~ Status
#   )) %>%
#   casebase::popTime(event = "Status", time = "ftime")
# 
# plot_grid(plot(pt_object) + ggtitle("Relapse"),
#           plot(pt_object_comp, point.colour = "blue") + ggtitle("Competing events"),
#           labels = c("A", "B"))

popTimeData <- popTime(data = bmtcrr, time = "ftime")

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through four different case studies.
  | *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, "\\proglang{R}", data visualization]
  plain:     [survival analysis, absolute risk, R, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Logistic regression had already been used in the context of discrete-time survival modeling [@cox1972regression]. But in the context of continuous-time survival modeling, using the framework of logistic regression opens the door to an extensive array of powerful modeling tools. Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function. Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.

@@

## Likelihood and estimating function

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, their sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base (e.g. the sum of all individual follow-up times). More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score. Non-uniform sampling mechanisms can increase the efficiency of the case-base estimators.

## Common parametric models

@@

## Competing-risk analysis

@@

## Variable selection

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

@@

$$ CI_j(t) = \int_0^t f_j(u)du.$$
Again, he integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration. This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.

@@

## Different functional forms of time

@@

@@

## Time-dependent hazard ratios

@@

In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

@@

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit_inter,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = TRUE)

axis(2, at = seq(0,1.50, by=0.25), las = 2)
axis(1, at = seq(min(new_time),max(new_time), by=1))
for( i in seq(0,1.50, by=0.25)) {
  abline(h = i, lty = 1, col = "lightgrey")
}
for( i in seq(min(new_time),max(new_time), by=1)) {
  abline(v = i, lty = 1, col = "lightgrey")
}
for(i in seq_along(seq(0, 0.75, by = 0.25))) {
  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
}
lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
```

@@

@@

In the next case study, we show how case-base sampling can be used in the context of a competing-risk analysis. For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.

@@

The data contains information on 177 patients who received a stem-cell transplant for acute leukemia. The event of interest is relapse, but other competing causes (e.g. death, progression, graft failure, graft-versus-host disease) were also recorded. Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.

First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B). In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}
popTimeData <- popTime(data = bmtcrr, time = "ftime")

plot(popTimeData, 
     add.competing.event = TRUE,
     comprisk = TRUE,
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     competing.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si())
```
